---
title: "Comparing CLAMPbase and CLAMPfull"
author: "Maria Chikina"
package: CLAMP
output:
  BiocStyle::html_document:
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Comparing CLAMPbase and CLAMPfull}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

  
```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    cache = FALSE,
    dev = "png",
    dev.args = list(type = "cairo")
)
library(CLAMP)
```

# Overview

This vignette compares the two main models in the **CLAMP** package:

1. **CLAMPbase**: Unsupervised matrix factorization for dimensionality reduction without prior knowledge.
2. **CLAMPfull**: Pathway-guided refinement that integrates biological prior knowledge to improve interpretability.

Using a small human whole blood RNA-Seq dataset, we demonstrate that incorporating pathway priors in `CLAMPfull` improves the biological interpretability of latent variables compared to the baseline `CLAMPbase` model.

We illustrate how to:

- Fit both `CLAMPbase` and `CLAMPfull` models,
- Compare their performance using cell type correlations, and
- Verify that the FBM (Filebacked Big Matrix) implementation produces identical results.


## 1. Load Data and Normalize

```{r load-data}
data("dataWholeBlood")
data("majorCellTypes")
data("celltypeTargets")

# Scale each gene to mean 0 and variance 1
dataWholeBlood <- tscale(dataWholeBlood)
```

## 2. Load Prior Knowledge Matrices

```{r load-priors-download, eval=FALSE}
# How to download pathway and cell marker libraries from Enrichr.
# Not run during vignette build to avoid network calls; pre-fetched
# .rds files are loaded in the next chunk instead.
gmtList <- list(
    CellMarkers = getGMT(
        "https://maayanlab.cloud/Enrichr/geneSetLibrary?mode=text&libraryName=CellMarker_2024",
        "CellMarker_2024"
    ),
    KEGG = getGMT(
        "https://maayanlab.cloud/Enrichr/geneSetLibrary?mode=text&libraryName=KEGG_2021_Human",
        "KEGG_2021_Human"
    )
)
```

```{r load-priors}
# Load pre-fetched gene set libraries bundled with the package
gmtList <- list(
    CellMarkers = readRDS(system.file("extdata", "CellMarker_2024.rds", package = "CLAMP")),
    KEGG        = readRDS(system.file("extdata", "KEGG_2021_Human.rds", package = "CLAMP"))
)

# Combine into a single sparse matrix
pathMatCell <- gmtListToSparseMat(gmtList)

# Load additional xCell reference matrix
data("xCell")

# Match pathways to the gene space of whole blood
matchedPathsWB <- getMatchedPathwayMatList(
    pathMatCell,
    xCell,
    new.genes = rownames(dataWholeBlood),
    min.genes = 2
)
```

## 3. Compute SVD and infer k

```{r compute-svd}
# Compute SVD (k = n_samples - 1)
set.seed(1)
wb_svd_k <- ncol(dataWholeBlood) - 1
dataWholeBlood.svd <- rsvd::rsvd(dataWholeBlood, k = wb_svd_k)

# Infer clamp_k from singular values
wb_clamp_k <- num.pc(list(d = dataWholeBlood.svd$d)) * 2
wb_clamp_k
```

## 4. Fit CLAMPbase and CLAMPfull

```{r fit-clampbase}
# Fit baseline CLAMP
wb.clamp.base <- CLAMPbase(
    dataWholeBlood,
    svdres = dataWholeBlood.svd,
    clamp_k = wb_clamp_k,
    trace = FALSE,
    adaptive.p = 0.05
)
```

```{r fit-clampfull}
# Fit variance-prior extended CLAMP model
wb.clamp.full <- CLAMPfull(
    dataWholeBlood,
    priorMat = matchedPathsWB,
    svdres = dataWholeBlood.svd,
    clamp.base.result = wb.clamp.base,
    clamp_k = wb_clamp_k,
    trace = TRUE,
    use_cpp = TRUE
)
```

## 5. Compare CLAMPbase vs CLAMPfull

This plot compares the maximum Spearman correlation for each major blood cell type between `CLAMPbase` and `CLAMPfull`.

Points above the red dashed line indicate improved correspondence when biological priors are included.

Most cell types show higher correlations under `CLAMPfull`, demonstrating that integrating pathway information helps capture more biologically meaningful latent variables.

```{r compare-models, fig.width=6, fig.height=5}
output <- compareBs(
    wb.clamp.base,
    wb.clamp.full,
    celltypeTargets,
    "CLAMPbase",
    "CLAMPfull",
    m = "s"
)

output$plot
```
## 7. Verify FBM Implementation

We verify that `CLAMPfull` produces identical results when using a file-backed matrix (FBM) input instead of an in-memory matrix.

```{r fit-clampfull-fbm}
dataWholeBloodFBM <- bigstatsr::as_FBM(dataWholeBlood)

wb.clamp.full.fbm <- CLAMPfull(
    dataWholeBloodFBM,
    priorMat = matchedPathsWB,
    svdres = dataWholeBlood.svd,
    clamp.base.result = wb.clamp.base,
    clamp_k = wb_clamp_k,
    trace = TRUE,
    use_cpp = TRUE
)
```

The FBM implementation produces identical results:

```{r compare-fbm, fig.width=6, fig.height=5}
output <- compareBs(
    wb.clamp.full,
    wb.clamp.full.fbm,
    celltypeTargets,
    "CLAMPbase",
    "CLAMPfull",
    m = "s"
)

output$plot
```

# Session Information

```{r session-info}
sessionInfo()
```
